---
title: "PIMA - Prediction Model"
author: "Arshad Ahamed Shajahan / 25242322"
date: "2026-03-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(25242322)
library(tidyverse)
library(ggplot2)
library(GGally)
library(mlbench)
library(mice)
library(VIM)
library(caret)
library(randomForest)
library(MASS)
library(gbm)
library(pROC)
library(knitr)
library(tree)
```

## PIMA Dataset

```{r}
# Loading the dataset and visualizing the summary

data("PimaIndiansDiabetes2")
pima_full <- PimaIndiansDiabetes2
pima_cc <- na.omit(PimaIndiansDiabetes2)
summary(pima_full)
dim(pima_full)
summary(pima_cc)
dim(pima_cc)
```

```{r}
# Look for missing data
md.pattern(pima_full)

na_summary <- colSums(is.na(pima_full))
print(na_summary)
```
We can notice that there are many missing values in the data. Variable like 'Insulin' has as high as 374 missing values.
Good thing to note is that, there are no missing values in our outcome variable 'diabetes'.

Since the percentage of missing value is very high, we can create and evaluate models by complete case analysis and also by imputing missing values and compare the results between them.


```{r}
# Distribution of outcome variable
pima_full %>%
  pivot_longer(cols = -diabetes, names_to = "Variable", values_to = "Value") %>%
  filter(!is.na(Value)) %>%
  ggplot(aes(x = Value, fill = diabetes)) +
  geom_density(alpha = 0.45) +
  facet_wrap(~Variable, scales = "free", ncol = 4) +
  scale_fill_manual(values = c("neg" = "#2A9D8F", "pos" = "#E63946")) +
  labs(title = "Predictor Distributions by Diabetes Status",
       x = NULL, y = "Density", fill = "Diabetes") +
  theme_bw() +
  theme(legend.position = "bottom")
```



```{r}
# Correlation matrix (complete cases only for computation)
ggpairs(pima_cc,
        aes(colour = diabetes, alpha = 0.4),
        upper = list(continuous = wrap("cor", size = 3)),
        lower = list(continuous = wrap("smooth", se = FALSE, size = 0.3)),
        diag  = list(continuous = wrap("densityDiag")),
        columns = 1:8) +
  labs(title = "Pairwise Relationships") +
  theme_bw(base_size = 9) +
  theme(legend.position = "bottom")
```

## Complete Case Analysis

```{r}
# Data Overview
cat("Original dataset: ", nrow(pima_full), "rows\n")
cat("Complete cases:   ", nrow(pima_cc),   "rows\n")
cat("Rows dropped:     ", nrow(pima_full) - nrow(pima_cc), 
    paste0("(", round((1 - nrow(pima_cc)/nrow(pima_full))*100, 1), "% lost)\n"))
```

```{r}
pima_full %>% mutate(Status = ifelse(complete.cases(.), "Complete", "Incomplete"))
```

## KNN Imputation

```{r}
pima_knn <- kNN(pima_full, k = 5, imp_var = FALSE)

cat("KNN Imputed dataset:", nrow(pima_knn), "rows (100% retained)\n")
cat("Missing values after imputation:", sum(is.na(pima_knn)), "\n")
summary(pima_knn)
```

## Data Splitting

```{r}
# Splitting the dataset 70-30 for training and testing
train_index <- createDataPartition(pima_full$diabetes, p = 0.7,list = FALSE)

# Split the full dataset first
pima_train_full <- pima_full[ train_index, ]
pima_test_full  <- pima_full[-train_index, ]

cat("Full dataset split:\n")
cat("Training set:", nrow(pima_train_full), "rows\n")
cat("Test set:", nrow(pima_test_full),  "rows\n")

# Train-Test split for Complete Case Analysis
pima_cc_train <- na.omit(pima_train_full)
pima_cc_test  <- na.omit(pima_test_full)

cat("\nCCA:\n")
cat("Training set:", nrow(pima_cc_train), "rows\n")
cat("Test set:", nrow(pima_cc_test),  "rows\n")
cat("Train rows dropped:", nrow(pima_train_full) - nrow(pima_cc_train), "\n")
cat("Test rows dropped: ", nrow(pima_test_full)  - nrow(pima_cc_test),  "\n")

# Train-Test split for KNN Imputation
pima_knn_train <- kNN(pima_train_full, k = 5, imp_var = FALSE)
pima_knn_test  <- kNN(pima_test_full,  k = 5, imp_var = FALSE)

cat("\nKNN:\n")
cat("Training set:", nrow(pima_knn_train),"\n")
cat("Test set:    ", nrow(pima_knn_test),"\n")
```

## Feature Selection

```{r}
# Backward Stepwise AIC for full and KNN imputed data

# Full Model
full_model_cca <- glm(diabetes ~ ., data = pima_cc_train, family = binomial)
step_model_cca <- stepAIC(full_model_cca, direction = "backward", trace = FALSE)

selected_cca <- names(coef(step_model_cca))[-1]
cat("CCA selected predictors:", paste(selected_cca, collapse = ", "), "\n")
cat("AIC for selected predictors:", round(AIC(step_model_cca), 2),
    "| Full:", round(AIC(full_model_cca), 2), "\n")

# KNN Imputed Model
full_model_knn <- glm(diabetes ~ ., data = pima_knn_train, family = binomial)
step_model_knn <- stepAIC(full_model_knn, direction = "backward", trace = FALSE)

selected_knn <- names(coef(step_model_knn))[-1]
cat("\nKNN Imputed selected predictors:", paste(selected_knn, collapse = ", "), "\n")
cat("AIC for selected predictors:", round(AIC(step_model_knn), 2),
    "| Full:", round(AIC(full_model_knn), 2), "\n")
```
```{r}
# Predictor Selection

predictors <- names(pima_full)[names(pima_full) != "diabetes"]

var_compare <- data.frame(
  Variable = predictors,
  CCA      = ifelse(predictors %in% selected_cca, "✅", "❌"),
  KNN      = ifelse(predictors %in% selected_knn, "✅", "❌")
)

knitr::kable(var_compare,
             caption = "Variables Selected by Backward Stepwise AIC (Training Set)")
```
```{r}
# Prepare the training and testing variables
pima_cc_train_sel <- pima_cc_train %>% dplyr::select(all_of(c(selected_cca, "diabetes")))
pima_cc_test_sel  <- pima_cc_test  %>% dplyr::select(all_of(c(selected_cca, "diabetes")))
pima_knn_train_sel <- pima_knn_train %>% dplyr::select(all_of(c(selected_knn, "diabetes")))
pima_knn_test_sel  <- pima_knn_test  %>% dplyr::select(all_of(c(selected_knn, "diabetes")))

ctrl <- trainControl(
  method = "cv",
  number = 10,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final")

```

## Model Fitting

```{r}
# Logistic Regression For CCA
lr_cca <- train(diabetes ~ ., data = pima_cc_train_sel, method = "glm", family = binomial,
  trControl = ctrl, metric = "ROC")
print(lr_cca)

# Random Forest For CCA
rf_cca <- train(diabetes ~ ., data = pima_cc_train_sel, method= "rf", trControl  = ctrl,
  metric = "ROC", tuneLength = 3)
print(rf_cca)

# Logistic Regression For KNN
lr_knn <- train(diabetes ~ ., data = pima_knn_train_sel, method = "glm", family = binomial,
  trControl = ctrl, metric = "ROC")
print(lr_cca)

# Random Forest For KNN
rf_knn <- train(diabetes ~ ., data = pima_knn_train_sel, method= "rf", trControl  = ctrl,
  metric = "ROC", tuneLength = 3)
print(rf_cca)
```

## Model - Training Set Evaluation

```{r}
# Training set evaluation
best_lr_cca <- lr_cca$results[which.max(lr_cca$results$ROC), ]
best_rf_cca <- rf_cca$results[which.max(rf_cca$results$ROC), ]
best_lr_knn <- lr_knn$results[which.max(lr_knn$results$ROC), ]
best_rf_knn <- rf_knn$results[which.max(rf_knn$results$ROC), ]

cv_results <- data.frame(
  Model = c(
    "Logistic Regression | CCA",
    "Random Forest       | CCA",
    "Logistic Regression | KNN",
    "Random Forest       | KNN"),
  AUC = round(c(best_lr_cca$ROC,  best_rf_cca$ROC,
                best_lr_knn$ROC,  best_rf_knn$ROC),  3),
  AUC_SD = round(c(best_lr_cca$ROCSD, best_rf_cca$ROCSD,
                   best_lr_knn$ROCSD, best_rf_knn$ROCSD), 3),
  Sensitivity = round(c(best_lr_cca$Sens, best_rf_cca$Sens,
                        best_lr_knn$Sens, best_rf_knn$Sens), 3),
  Specificity = round(c(best_lr_cca$Spec, best_rf_cca$Spec,
                        best_lr_knn$Spec, best_rf_knn$Spec), 3))

knitr::kable(cv_results,
             col.names = c("Model", "AUC", "AUC SD",
                           "Sensitivity", "Specificity"),
             caption   = "10-Fold CV Performance (Training Set)")
```


## Prediction and Confusion Matrix

```{r}
# Logistic Regression For CCA
pred_class_lr_cca <- predict(lr_cca,
                              newdata = pima_cc_test_sel,
                              type    = "raw")
pred_prob_lr_cca  <- predict(lr_cca,
                              newdata = pima_cc_test_sel,
                              type    = "prob")
cm_lr_cca <- confusionMatrix(
  data      = factor(pred_class_lr_cca,         levels = c("neg", "pos")),
  reference = factor(pima_cc_test_sel$diabetes, levels = c("neg", "pos")),
  positive  = "pos")
print(cm_lr_cca)


# Random Forest For CCA
pred_class_rf_cca <- predict(rf_cca,
                              newdata = pima_cc_test_sel,
                              type    = "raw")
pred_prob_rf_cca  <- predict(rf_cca,
                              newdata = pima_cc_test_sel,
                              type    = "prob")
cm_rf_cca <- confusionMatrix(
  data      = factor(pred_class_rf_cca,         levels = c("neg", "pos")),
  reference = factor(pima_cc_test_sel$diabetes, levels = c("neg", "pos")),
  positive  = "pos")
print(cm_rf_cca)


# Logistic Regression For KNN
pred_class_lr_knn <- predict(lr_knn,
                              newdata = pima_knn_test_sel,
                              type    = "raw")
pred_prob_lr_knn  <- predict(lr_knn,
                              newdata = pima_knn_test_sel,
                              type    = "prob")
cm_lr_knn <- confusionMatrix(
  data      = factor(pred_class_lr_knn,          levels = c("neg", "pos")),
  reference = factor(pima_knn_test_sel$diabetes, levels = c("neg", "pos")),
  positive  = "pos")
print(cm_lr_knn)



# Random Forest For KNN
pred_class_rf_knn <- predict(rf_knn,
                              newdata = pima_knn_test_sel,
                              type    = "raw")
pred_prob_rf_knn  <- predict(rf_knn,
                              newdata = pima_knn_test_sel,
                              type    = "prob")
cm_rf_knn <- confusionMatrix(
  data      = factor(pred_class_rf_knn,          levels = c("neg", "pos")),
  reference = factor(pima_knn_test_sel$diabetes, levels = c("neg", "pos")),
  positive  = "pos")
print(cm_rf_knn)
```

## ROC Analysis

```{r}

roc_lr_cca_test <- roc(pima_cc_test_sel$diabetes,  pred_prob_lr_cca$pos,
                       levels = c("neg","pos"), direction = "<", quiet = TRUE)

roc_rf_cca_test <- roc(pima_cc_test_sel$diabetes,  pred_prob_rf_cca$pos,
                       levels = c("neg","pos"), direction = "<", quiet = TRUE)

roc_lr_knn_test <- roc(pima_knn_test_sel$diabetes, pred_prob_lr_knn$pos,
                       levels = c("neg","pos"), direction = "<", quiet = TRUE)

roc_rf_knn_test <- roc(pima_knn_test_sel$diabetes, pred_prob_rf_knn$pos,
                       levels = c("neg","pos"), direction = "<", quiet = TRUE)
```

```{r}
roc_test_df <- bind_rows(
  data.frame(
    FPR   = 1 - roc_lr_cca_test$specificities,
    TPR   = roc_lr_cca_test$sensitivities,
    Model = paste0("LR  | CCA  (AUC=", round(auc(roc_lr_cca_test), 3), ")")
  ),
  data.frame(
    FPR   = 1 - roc_rf_cca_test$specificities,
    TPR   = roc_rf_cca_test$sensitivities,
    Model = paste0("RF  | CCA  (AUC=", round(auc(roc_rf_cca_test), 3), ")")
  ),
  data.frame(
    FPR   = 1 - roc_lr_knn_test$specificities,
    TPR   = roc_lr_knn_test$sensitivities,
    Model = paste0("LR  | KNN  (AUC=", round(auc(roc_lr_knn_test), 3), ")")
  ),
  data.frame(
    FPR   = 1 - roc_rf_knn_test$specificities,
    TPR   = roc_rf_knn_test$sensitivities,
    Model = paste0("RF  | KNN  (AUC=", round(auc(roc_rf_knn_test), 3), ")")
  )
)

ggplot(roc_test_df, aes(x = FPR, y = TPR, colour = Model)) +
  geom_line(linewidth = 0.9) +
  geom_abline(linetype = "dashed", colour = "grey50") +
  scale_colour_manual(values = c("#E63946","#457B9D","#F4A261","#2A9D8F")) +
  labs(title    = "ROC Curves (30% Held Out Test Set)",
       subtitle = "All four Model × Pre-processing Combinations",
       x        = "False Positive Rate (1 – Specificity)",
       y        = "True Positive Rate (Sensitivity)",
       colour   = NULL) +
  theme_bw(base_size = 13) +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold"))

```

## Model Summary

```{r}
## Combined Performance Summary — CV vs Test Set

test_results <- data.frame(
  Model = c(
    "Logistic Regression | CCA",
    "Random Forest       | CCA",
    "Logistic Regression | KNN",
    "Random Forest       | KNN"
  ),
  CV_AUC = round(c(best_lr_cca$ROC, best_rf_cca$ROC,
                   best_lr_knn$ROC, best_rf_knn$ROC), 3),
  Test_AUC = round(c(auc(roc_lr_cca_test), auc(roc_rf_cca_test),
                     auc(roc_lr_knn_test), auc(roc_rf_knn_test)), 3),
  Test_Accuracy = round(c(cm_lr_cca$overall["Accuracy"],
                          cm_rf_cca$overall["Accuracy"],
                          cm_lr_knn$overall["Accuracy"],
                          cm_rf_knn$overall["Accuracy"]), 3),
  Test_Sensitivity = round(c(cm_lr_cca$byClass["Sensitivity"],
                             cm_rf_cca$byClass["Sensitivity"],
                             cm_lr_knn$byClass["Sensitivity"],
                             cm_rf_knn$byClass["Sensitivity"]), 3),
  Test_Specificity = round(c(cm_lr_cca$byClass["Specificity"],
                             cm_rf_cca$byClass["Specificity"],
                             cm_lr_knn$byClass["Specificity"],
                             cm_rf_knn$byClass["Specificity"]), 3),
  Test_PPV = round(c(cm_lr_cca$byClass["Pos Pred Value"],
                     cm_rf_cca$byClass["Pos Pred Value"],
                     cm_lr_knn$byClass["Pos Pred Value"],
                     cm_rf_knn$byClass["Pos Pred Value"]), 3),
  Test_NPV = round(c(cm_lr_cca$byClass["Neg Pred Value"],
                     cm_rf_cca$byClass["Neg Pred Value"],
                     cm_lr_knn$byClass["Neg Pred Value"],
                     cm_rf_knn$byClass["Neg Pred Value"]), 3)
)

knitr::kable(test_results,
             col.names = c("Model", "CV AUC", "Test AUC", "Accuracy",
                           "Sensitivity", "Specificity", "PPV", "NPV"),
             caption   = "CV vs Test Set Performance — All Models")


```

```{r}

# Visualise CV vs Test AUC side by side
test_results %>%
  dplyr::select(Model, CV_AUC, Test_AUC) %>%
  pivot_longer(cols = c(CV_AUC, Test_AUC),
               names_to  = "Phase",
               values_to = "AUC") %>%
  mutate(Phase = recode(Phase,
                        "CV_AUC"   = "10-Fold CV (Train)",
                        "Test_AUC" = "Hold-Out Test")) %>%
  ggplot(aes(x = Model, y = AUC, fill = Phase)) +
  geom_col(position = "dodge", width = 0.6) +
  geom_text(aes(label = round(AUC, 3)),
            position = position_dodge(width = 0.6),
            vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = c("10-Fold CV (Train)" = "#457B9D",
                               "Hold-Out Test"      = "#E63946")) +
  scale_y_continuous(limits = c(0, 1.05)) +
  labs(title    = "CV vs Test Set AUC — Overfitting Check",
       subtitle = "A large drop from CV to Test indicates overfitting",
       x = NULL, y = "AUC", fill = NULL) +
  theme_bw(base_size = 12) +
  theme(legend.position  = "bottom",
        plot.title       = element_text(face = "bold"),
        axis.text.x      = element_text(angle = 15, hjust = 1))
```

```{r}
# Variable Importance
vi_df <- varImp(rf_knn)$importance %>%
  tibble::rownames_to_column("Variable") %>%
  rename(Importance = Overall) %>%
  arrange(desc(Importance))

knitr::kable(vi_df, digits = 2,
             caption = "Variable Importance — Random Forest | KNN (Training Set)")

ggplot(vi_df, aes(x = reorder(Variable, Importance),
                  y = Importance, fill = Importance)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_gradient(low = "#A8DADC", high = "#1D3557") +
  labs(title    = "Variable Importance — Random Forest | KNN",
       subtitle = "Mean Decrease in Gini (10-Fold CV on Training Set)",
       x = NULL, y = "Importance") +
  theme_bw(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

```{r}
# Identify best model by Test AUC
best_row <- test_results[which.max(test_results$Test_AUC), ]

cat("Best Model by Test AUC:\n")
cat("  Model       :", best_row$Model,"\n")
cat("  CV AUC      :", best_row$CV_AUC,"\n")
cat("  Test AUC    :", best_row$Test_AUC,"\n")
cat("  Accuracy    :", best_row$Test_Accuracy,"\n")
cat("  Sensitivity :", best_row$Test_Sensitivity,"\n")
cat("  Specificity :", best_row$Test_Specificity, "\n")
cat("  PPV         :", best_row$Test_PPV,"\n")
cat("  NPV         :", best_row$Test_NPV,"\n")

```

## Further Analysis on Trees

```{r}
## Fitting Trees on CCA

tree_cca <- tree(diabetes ~ ., data = pima_cc_train_sel)
summary(tree_cca)
plot(tree_cca)
text(tree_cca, pretty = 0)

# Prediction
tree_pred_cca <- predict(tree_cca, newdata = pima_cc_test_sel, type = "class")
tree_prob_cca <- predict(tree_cca, newdata = pima_cc_test_sel, type = "vector")

cm_tree_cca <- confusionMatrix(
  data      = factor(tree_pred_cca,             levels = c("neg", "pos")),
  reference = factor(pima_cc_test_sel$diabetes, levels = c("neg", "pos")),
  positive  = "pos"
)
print(cm_tree_cca)
```



```{r}
# Fitting Trees on KNN Imputed data

tree_knn <- tree(diabetes ~ ., data = pima_knn_train_sel)
summary(tree_knn)
plot(tree_knn)
text(tree_knn, pretty = 0)

# Prediction
tree_pred_knn <- predict(tree_knn, newdata = pima_knn_test_sel, type = "class")
tree_prob_knn <- predict(tree_knn, newdata = pima_knn_test_sel, type = "vector")

cm_tree_knn <- confusionMatrix(
  data      = factor(tree_pred_knn,              levels = c("neg", "pos")),
  reference = factor(pima_knn_test_sel$diabetes, levels = c("neg", "pos")),
  positive  = "pos"
)
print(cm_tree_knn)
```

## Trees - Pruning

```{r}
# Pruning for CCA
cv_tree_cca <- cv.tree(tree_cca, FUN = prune.misclass)

par(mfrow = c(1, 2))
plot(cv_tree_cca$size, cv_tree_cca$dev, type = "b")
plot(cv_tree_cca$k, cv_tree_cca$dev, type = "b")
par(mfrow = c(1, 1))  

best_size_cca <- cv_tree_cca$size[which.min(cv_tree_cca$dev)]
prune_cca <- prune.misclass(tree_cca, best = best_size_cca)

plot(prune_cca)
text(prune_cca, pretty = 0)

# Predcition
prune_pred_cca <- predict(prune_cca, newdata = pima_cc_test_sel, type = "class")
prune_prob_cca <- predict(prune_cca, newdata = pima_cc_test_sel, type = "vector")

cm_prune_cca <- confusionMatrix(
  data = factor(prune_pred_cca, levels = c("neg", "pos")),
  reference = factor(pima_cc_test_sel$diabetes, levels = c("neg", "pos")),
  positive = "pos")
print(cm_prune_cca)

# Result
cat("Full tree error rate: ",
    round(mean(tree_pred_cca  != pima_cc_test_sel$diabetes), 3), "\n")
cat("Pruned tree error rate: ",
    round(mean(prune_pred_cca != pima_cc_test_sel$diabetes), 3), "\n")

```


```{r}
# Pruning for KNN

cv_tree_knn <- cv.tree(tree_knn, FUN = prune.misclass)

par(mfrow = c(1, 2))
plot(cv_tree_knn$size, cv_tree_knn$dev, type = "b")
plot(cv_tree_knn$k, cv_tree_knn$dev, type = "b")
par(mfrow = c(1, 1))

best_size_knn <- cv_tree_knn$size[which.min(cv_tree_knn$dev)]
prune_knn <- prune.misclass(tree_knn, best = best_size_knn)

plot(prune_knn)
text(prune_knn, pretty = 0)

# Prediction
prune_pred_knn <- predict(prune_knn, newdata = pima_knn_test_sel, type = "class")
prune_prob_knn <- predict(prune_knn, newdata = pima_knn_test_sel, type = "vector")

cm_prune_knn <- confusionMatrix(
  data = factor(prune_pred_knn,             levels = c("neg", "pos")),
  reference = factor(pima_knn_test_sel$diabetes, levels = c("neg", "pos")),
  positive = "pos")
print(cm_prune_knn)

# Result
cat("Full tree error rate:   ",
    round(mean(tree_pred_knn  != pima_knn_test_sel$diabetes), 3), "\n")
cat("Pruned tree error rate: ",
    round(mean(prune_pred_knn != pima_knn_test_sel$diabetes), 3), "\n")

```

## Trees - Bagging

```{r}
# Bagging for CCA

n_pred_cca <- ncol(pima_cc_train_sel) - 1

bag_cca <- randomForest(
  diabetes ~ .,
  data       = pima_cc_train_sel,
  mtry       = n_pred_cca,
  ntree      = 500,
  importance = TRUE)
print(bag_cca)

# Prediction
bag_pred_cca <- predict(bag_cca, newdata = pima_cc_test_sel, type = "class")
bag_prob_cca <- predict(bag_cca, newdata = pima_cc_test_sel, type = "prob")

cm_bag_cca <- confusionMatrix(
  data      = factor(bag_pred_cca,              levels = c("neg", "pos")),
  reference = factor(pima_cc_test_sel$diabetes, levels = c("neg", "pos")),
  positive  = "pos")
print(cm_bag_cca)

varImpPlot(bag_cca, main = "Variable Importance — Bagging | CCA")
```



```{r}
# Bagging for KNN

n_pred_knn <- ncol(pima_knn_train_sel) - 1

bag_knn <- randomForest(
  diabetes ~ .,
  data       = pima_knn_train_sel,
  mtry       = n_pred_knn,
  ntree      = 500,
  importance = TRUE)
print(bag_knn)

# PRediction
bag_pred_knn <- predict(bag_knn, newdata = pima_knn_test_sel, type = "class")
bag_prob_knn <- predict(bag_knn, newdata = pima_knn_test_sel, type = "prob")

cm_bag_knn <- confusionMatrix(
  data      = factor(bag_pred_knn,               levels = c("neg", "pos")),
  reference = factor(pima_knn_test_sel$diabetes, levels = c("neg", "pos")),
  positive  = "pos")
print(cm_bag_knn)

varImpPlot(bag_knn, main = "Variable Importance — Bagging | KNN")
```

## Trees - Boosting

```{r}
# Boosting for CCA

pima_cc_train_gbm <- pima_cc_train_sel %>%
  mutate(diabetes = ifelse(diabetes == "pos", 1, 0))

pima_cc_test_gbm <- pima_cc_test_sel %>%
  mutate(diabetes = ifelse(diabetes == "pos", 1, 0))

boost_cca <- gbm(
  diabetes ~ .,
  data             = pima_cc_train_gbm,
  distribution     = "bernoulli",
  n.trees          = 5000,
  interaction.depth = 2,
  shrinkage        = 0.001,
  cv.folds         = 10,
  verbose          = FALSE)
best_iter_cca <- gbm.perf(boost_cca, method = "cv", plot.it = FALSE)
cat("Optimal number of trees (CCA):", best_iter_cca, "\n")
summary(boost_cca, n.trees = best_iter_cca,
        main = "Relative Influence — Boosting | CCA")

# Prediction

boost_prob_cca <- predict(boost_cca,
                          newdata = pima_cc_test_gbm,
                          n.trees = best_iter_cca,
                          type    = "response")

boost_pred_cca <- factor(ifelse(boost_prob_cca >= 0.5, "pos", "neg"),
                         levels = c("neg", "pos"))


cm_boost_cca <- confusionMatrix(
  data      = boost_pred_cca,
  reference = factor(pima_cc_test_sel$diabetes, levels = c("neg", "pos")),
  positive  = "pos")
print(cm_boost_cca)
```

```{r}
# Boosting for KNN

pima_knn_train_gbm <- pima_knn_train_sel %>%
  mutate(diabetes = ifelse(diabetes == "pos", 1, 0))

pima_knn_test_gbm <- pima_knn_test_sel %>%
  mutate(diabetes = ifelse(diabetes == "pos", 1, 0))

boost_knn <- gbm(
  diabetes ~ .,
  data              = pima_knn_train_gbm,
  distribution      = "bernoulli",
  n.trees           = 5000,
  interaction.depth = 2,
  shrinkage         = 0.001,
  cv.folds          = 10,
  verbose           = FALSE
)

best_iter_knn <- gbm.perf(boost_knn, method = "cv", plot.it = FALSE)
cat("Optimal number of trees (KNN):", best_iter_knn, "\n")
summary(boost_knn, n.trees = best_iter_knn,
        main = "Relative Influence — Boosting | KNN")

# Prediction
boost_prob_knn <- predict(boost_knn,
                          newdata = pima_knn_test_gbm,
                          n.trees = best_iter_knn,
                          type    = "response")

boost_pred_knn <- factor(ifelse(boost_prob_knn >= 0.5, "pos", "neg"),
                         levels = c("neg", "pos"))


cm_boost_knn <- confusionMatrix(
  data      = boost_pred_knn,
  reference = factor(pima_knn_test_sel$diabetes, levels = c("neg", "pos")),
  positive  = "pos"
)
print(cm_boost_knn)
```

## ROC Curve Analysis

```{r}
# ROC Curve

roc_tree_cca   <- roc(pima_cc_test_sel$diabetes,  prune_prob_cca[, "pos"],
                      levels = c("neg","pos"), direction = "<", quiet = TRUE)
roc_tree_knn   <- roc(pima_knn_test_sel$diabetes, prune_prob_knn[, "pos"],
                      levels = c("neg","pos"), direction = "<", quiet = TRUE)
roc_bag_cca    <- roc(pima_cc_test_sel$diabetes,  bag_prob_cca[, "pos"],
                      levels = c("neg","pos"), direction = "<", quiet = TRUE)
roc_bag_knn    <- roc(pima_knn_test_sel$diabetes, bag_prob_knn[, "pos"],
                      levels = c("neg","pos"), direction = "<", quiet = TRUE)
roc_boost_cca  <- roc(pima_cc_test_sel$diabetes,  boost_prob_cca,
                      levels = c("neg","pos"), direction = "<", quiet = TRUE)
roc_boost_knn  <- roc(pima_knn_test_sel$diabetes, boost_prob_knn,
                      levels = c("neg","pos"), direction = "<", quiet = TRUE)

# Combine all models into one data frame
roc_all_df <- bind_rows(
  data.frame(FPR = 1 - roc_lr_cca_test$specificities,
             TPR = roc_lr_cca_test$sensitivities,
             Model = paste0("LR      | CCA (AUC=", round(auc(roc_lr_cca_test),  3), ")")),
  data.frame(FPR = 1 - roc_lr_knn_test$specificities,
             TPR = roc_lr_knn_test$sensitivities,
             Model = paste0("LR      | KNN (AUC=", round(auc(roc_lr_knn_test),  3), ")")),
  data.frame(FPR = 1 - roc_rf_cca_test$specificities,
             TPR = roc_rf_cca_test$sensitivities,
             Model = paste0("RF      | CCA (AUC=", round(auc(roc_rf_cca_test),  3), ")")),
  data.frame(FPR = 1 - roc_rf_knn_test$specificities,
             TPR = roc_rf_knn_test$sensitivities,
             Model = paste0("RF      | KNN (AUC=", round(auc(roc_rf_knn_test),  3), ")")),
  data.frame(FPR = 1 - roc_tree_cca$specificities,
             TPR = roc_tree_cca$sensitivities,
             Model = paste0("Tree    | CCA (AUC=", round(auc(roc_tree_cca),     3), ")")),
  data.frame(FPR = 1 - roc_tree_knn$specificities,
             TPR = roc_tree_knn$sensitivities,
             Model = paste0("Tree    | KNN (AUC=", round(auc(roc_tree_knn),     3), ")")),
  data.frame(FPR = 1 - roc_bag_cca$specificities,
             TPR = roc_bag_cca$sensitivities,
             Model = paste0("Bagging | CCA (AUC=", round(auc(roc_bag_cca),      3), ")")),
  data.frame(FPR = 1 - roc_bag_knn$specificities,
             TPR = roc_bag_knn$sensitivities,
             Model = paste0("Bagging | KNN (AUC=", round(auc(roc_bag_knn),      3), ")")),
  data.frame(FPR = 1 - roc_boost_cca$specificities,
             TPR = roc_boost_cca$sensitivities,
             Model = paste0("Boost   | CCA (AUC=", round(auc(roc_boost_cca),    3), ")")),
  data.frame(FPR = 1 - roc_boost_knn$specificities,
             TPR = roc_boost_knn$sensitivities,
             Model = paste0("Boost   | KNN (AUC=", round(auc(roc_boost_knn),    3), ")")))

ggplot(roc_all_df, aes(x = FPR, y = TPR, colour = Model)) +
  geom_line(linewidth = 0.8) +
  geom_abline(linetype = "dashed", colour = "grey50") +
  labs(title    = "ROC Curves — All Models × All Strategies",
       subtitle = "30% Held-Out Test Set",
       x        = "False Positive Rate (1 – Specificity)",
       y        = "True Positive Rate (Sensitivity)",
       colour   = NULL) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom",
        legend.text     = element_text(size = 8),
        plot.title      = element_text(face = "bold")) +
  guides(colour = guide_legend(ncol = 2))

```

## Consolidated Results For All Models

```{r}
all_results <- data.frame(
  Model = c(
    "Logistic Regression | CCA",
    "Logistic Regression | KNN",
    "Random Forest       | CCA",
    "Random Forest       | KNN",
    "Pruned Tree         | CCA",
    "Pruned Tree         | KNN",
    "Bagging             | CCA",
    "Bagging             | KNN",
    "Boosting            | CCA",
    "Boosting            | KNN"),
  AUC = round(c(
    auc(roc_lr_cca_test),  auc(roc_lr_knn_test),
    auc(roc_rf_cca_test),  auc(roc_rf_knn_test),
    auc(roc_tree_cca),     auc(roc_tree_knn),
    auc(roc_bag_cca),      auc(roc_bag_knn),
    auc(roc_boost_cca),    auc(roc_boost_knn)), 3),
  Accuracy = round(c(
    cm_lr_cca$overall["Accuracy"],    cm_lr_knn$overall["Accuracy"],
    cm_rf_cca$overall["Accuracy"],    cm_rf_knn$overall["Accuracy"],
    cm_prune_cca$overall["Accuracy"], cm_prune_knn$overall["Accuracy"],
    cm_bag_cca$overall["Accuracy"],   cm_bag_knn$overall["Accuracy"],
    cm_boost_cca$overall["Accuracy"], cm_boost_knn$overall["Accuracy"]), 3),
  Sensitivity = round(c(
    cm_lr_cca$byClass["Sensitivity"],    cm_lr_knn$byClass["Sensitivity"],
    cm_rf_cca$byClass["Sensitivity"],    cm_rf_knn$byClass["Sensitivity"],
    cm_prune_cca$byClass["Sensitivity"], cm_prune_knn$byClass["Sensitivity"],
    cm_bag_cca$byClass["Sensitivity"],   cm_bag_knn$byClass["Sensitivity"],
    cm_boost_cca$byClass["Sensitivity"], cm_boost_knn$byClass["Sensitivity"]), 3),
  Specificity = round(c(
    cm_lr_cca$byClass["Specificity"],    cm_lr_knn$byClass["Specificity"],
    cm_rf_cca$byClass["Specificity"],    cm_rf_knn$byClass["Specificity"],
    cm_prune_cca$byClass["Specificity"], cm_prune_knn$byClass["Specificity"],
    cm_bag_cca$byClass["Specificity"],   cm_bag_knn$byClass["Specificity"],
    cm_boost_cca$byClass["Specificity"], cm_boost_knn$byClass["Specificity"]), 3),
  PPV = round(c(
    cm_lr_cca$byClass["Pos Pred Value"],    cm_lr_knn$byClass["Pos Pred Value"],
    cm_rf_cca$byClass["Pos Pred Value"],    cm_rf_knn$byClass["Pos Pred Value"],
    cm_prune_cca$byClass["Pos Pred Value"], cm_prune_knn$byClass["Pos Pred Value"],
    cm_bag_cca$byClass["Pos Pred Value"],   cm_bag_knn$byClass["Pos Pred Value"],
    cm_boost_cca$byClass["Pos Pred Value"], cm_boost_knn$byClass["Pos Pred Value"]), 3),
  NPV = round(c(
    cm_lr_cca$byClass["Neg Pred Value"],    cm_lr_knn$byClass["Neg Pred Value"],
    cm_rf_cca$byClass["Neg Pred Value"],    cm_rf_knn$byClass["Neg Pred Value"],
    cm_prune_cca$byClass["Neg Pred Value"], cm_prune_knn$byClass["Neg Pred Value"],
    cm_bag_cca$byClass["Neg Pred Value"],   cm_bag_knn$byClass["Neg Pred Value"],
    cm_boost_cca$byClass["Neg Pred Value"], cm_boost_knn$byClass["Neg Pred Value"] ), 3))

knitr::kable(all_results,
             col.names = c("Model", "AUC", "Accuracy",
                           "Sensitivity", "Specificity", "PPV", "NPV"),
             caption   = "Test Set Performance — All Models × All Strategies")
```

```{r}
all_results %>%
  arrange(desc(AUC)) %>%
  mutate(Model = factor(Model, levels = Model)) %>%
  ggplot(aes(x = Model, y = AUC, fill = AUC)) +
  geom_col(show.legend = FALSE, width = 0.7) +
  geom_text(aes(label = round(AUC, 3)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_fill_gradient(low = "#A8DADC", high = "#1D3557") +
  scale_y_continuous(limits = c(0, 1.05)) +
  labs(title    = "Test Set AUC — All Models Ranked",
       subtitle = "AUC Higher the Better",
       x = NULL, y = "AUC") +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

```